---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
a <- read.csv("./activity.csv", header = TRUE)
head(a)
str(a)
```

Then split the data frame into convenient variables representing each name

```{r, echo=TRUE}
steps <- a$steps
dates <- a$date
interval <- a$interval
```

## What is mean total number of steps taken per day?
First determine the total number of steps per day; placed in a variable sumday

```{r, echo=TRUE}
library(dplyr)
timedat <- group_by(a, date)
sumday <- summarise(timedat, totalsteps = sum(steps, na.rm=TRUE))
hist(sumday$totalsteps, breaks = 9, main = "Histogram of Total Steps Per Day", xlab = "Total Steps Taken")
meantot <- mean(sumday$totalsteps)
medtot <- median(sumday$totalsteps)
```

The mean total number of steps taken per day is `r meantot`
The median total number of steps taken per day is `r medtot`

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
